---
title: "07a_Functional_Analysis_GSEA (fixed)"
author: "automated-patch"
date: "2026-03-05"
output: pdf_document
---

# Functional Analysis/Network Analysis - Summary of the Workflow

1.  **Annotation:** Maps SOMAmers to Entrez IDs.
2.  **Ranking:** Uses WGCNA kME values to rank every protein in the dataset for each module.
3.  **Enrichment:** Runs `fgseaMultilevel` against three distinct databases (Hallmark, Reactome, GO).
4.  **Integration:** Combines results into a single, structured data object.
5.  **Visualization:** Creates heatmaps, dot plots, and network plots for interpretation.

# 1. Setup

```{r setup, include=FALSE}
library(here)
base_dir <- here::here("Proteomics")
knitr::opts_knit$set(root.dir = base_dir)

results_dir <- file.path(base_dir, "results", "functional_analysis_gsea")
plots_dir <- file.path(results_dir, "plots")
dir.create(plots_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r libs, message=FALSE, warning=FALSE}
pkgs <- c("WGCNA", "dplyr", "readr", "clusterProfiler", "org.Hs.eg.db", "ReactomePA",
          "ggplot2", "openxlsx", "fgsea", "msigdbr", "tidyr", "tibble", "pheatmap", "stringr", "enrichplot")

# Install missing packages
for(pk in pkgs){
  if(!requireNamespace(pk, quietly = TRUE)){
    message("Installing missing package: ", pk)
    is_bioc <- pk %in% c("WGCNA", "clusterProfiler", "org.Hs.eg.db", "ReactomePA", "enrichplot", "fgsea", "msigdbr")
    if(is_bioc){
      if(!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
      BiocManager::install(pk, ask = FALSE, update = FALSE)
    } else {
      install.packages(pk)
    }
  }
}

# Load all packages
lapply(pkgs, library, character.only = TRUE)
options(stringsAsFactors = FALSE)
```

# 2. Load Data & Prepare for GSEA

```{r load-data}
# Load WGCNA results (expr, moduleColors, MEs)
wgcna_file <- file.path(base_dir, "results", "wgcna", "WGCNA2_core_outputs.RData")
if(!file.exists(wgcna_file)) stop("WGCNA output file not found: ", wgcna_file)
load(wgcna_file)

# Validate loaded objects
req_objs <- c("expr", "moduleColors", "MEs")
if(!all(req_objs %in% ls())) stop("Loaded RData is missing one of: ", paste(req_objs, collapse=", "))
if(is.null(names(moduleColors))) names(moduleColors) <- colnames(expr)
stopifnot(identical(colnames(expr), names(moduleColors)))

# Get module list (excluding grey)
modules <- setdiff(unique(moduleColors), "grey")

# Compute kME (module membership)
datKME <- as.data.frame(cor(expr, MEs, use = "p"))
colnames(datKME) <- paste0("kME_", sub("^ME", "", colnames(MEs)))
rownames(datKME) <- colnames(expr)
```

# 3. Annotation Mapping (SOMAmer -> Entrez)

```{r id-mapping}
# Load feature annotation
analyte_file <- file.path(base_dir, "data", "CSF_SOMAscan7k_analyte_information.tsv")
if(!file.exists(analyte_file)) stop("Analyte info file not found: ", analyte_file)
analyte_info <- read.delim(analyte_file, sep = "\t", header = TRUE, check.names = FALSE)

# Create a mapping key from SOMAmer to Symbol
feature_key <- analyte_info %>%
  filter(Analytes %in% colnames(expr)) %>%
  select(Analytes, EntrezGeneSymbol) %>%
  distinct() %>%
  rename(SOMAmer = Analytes, SYMBOL = EntrezGeneSymbol)

# Map Symbol to Entrez ID
symbol_map <- bitr(
  feature_key$SYMBOL,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>% distinct(SYMBOL, .keep_all = TRUE)

# Final mapping table for the entire dataset
feature_map <- inner_join(feature_key, symbol_map, by = "SYMBOL")
message("Mapped ", nrow(feature_map), " SOMAmers to Entrez IDs.")
```

# 4. Load Gene Sets

```{r load-gene-sets}
hm <- msigdbr(species = "Homo sapiens", category = "H")
react <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
go <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")

gene_sets <- list(
  Hallmark = split(hm$entrez_gene, as.character(hm$gs_name)),
  Reactome = split(react$entrez_gene, as.character(react$gs_name)),
  GO_BP = split(go$entrez_gene, as.character(go$gs_name))
)

message("Loaded gene sets:")
message("  Hallmark: ", length(gene_sets$Hallmark))
message("  Reactome: ", length(gene_sets$Reactome))
message("  GO_BP: ", length(gene_sets$GO_BP))
```

# 5. GSEA Function

```{r gsea-function}
# Corrected GSEA function: Ranks ALL proteins by kME for a given module.
run_fgsea_module <- function(module, pathway_sets) {
  kme_col <- paste0("kME_", module)
  if (!kme_col %in% colnames(datKME)) return(NULL)

  # Rank all proteins by their kME for this module
  ranks_df <- data.frame(
    SOMAmer = rownames(datKME),
    kME = datKME[[kme_col]]
  ) %>%
    inner_join(feature_map, by = "SOMAmer") %>%
    filter(is.finite(kME)) %>%
    distinct(ENTREZID, .keep_all = TRUE)

  if (nrow(ranks_df) < 50) return(NULL)

  gene_ranks <- setNames(ranks_df$kME, ranks_df$ENTREZID)
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)

  fgseaMultilevel(
    pathways = pathway_sets,
    stats = gene_ranks,
    minSize = 15,
    maxSize = 500,
    nproc = 1 # Use 1 for safety in RMarkdown
  )
}
```

# 6. Run GSEA Across All Modules and Databases

```{r run-gsea-all}
all_gsea_results <- list()

for (mod in modules) {
  message("Processing module: ", mod)
  
  # Run GSEA for each database
  h_res <- run_fgsea_module(mod, gene_sets$Hallmark)
  r_res <- run_fgsea_module(mod, gene_sets$Reactome)
  g_res <- run_fgsea_module(mod, gene_sets$GO_BP)
  
  # Combine results for the module
  combined <- bind_rows(
    if (!is.null(h_res)) as_tibble(h_res) %>% mutate(Source = "Hallmark"),
    if (!is.null(r_res)) as_tibble(r_res) %>% mutate(Source = "Reactome"),
    if (!is.null(g_res)) as_tibble(g_res) %>% mutate(Source = "GO_BP")
  )
  
  if (nrow(combined) > 0) {
    all_gsea_results[[mod]] <- combined
  } else {
    message("  -> No significant GSEA results for module ", mod)
  }
}
```

# 7. Save All Results to Excel

```{r save-excel}
# Single combined file
wb_all <- createWorkbook()
for (mod in names(all_gsea_results)) {
  df <- all_gsea_results[[mod]] %>% filter(padj < 0.05) %>% arrange(Source, padj)
  if (nrow(df) > 0) {
    addWorksheet(wb_all, mod)
    writeData(wb_all, mod, df)
  }
}
saveWorkbook(wb_all, file.path(results_dir, "Combined_GSEA_Results.xlsx"), overwrite = TRUE)

# Database-specific files
for (src in names(gene_sets)) {
  wb_src <- createWorkbook()
  for (mod in names(all_gsea_results)) {
    df <- all_gsea_results[[mod]] %>% filter(Source == src, padj < 0.05) %>% arrange(padj)
    if (nrow(df) > 0) {
      addWorksheet(wb_src, mod)
      writeData(wb_src, mod, df)
    }
  }
  if(length(wb_src$worksheets) > 0){
    saveWorkbook(wb_src, file.path(results_dir, paste0("GSEA_", src, "_Results.xlsx")), overwrite = TRUE)
  }
}
message("GSEA results saved to: ", results_dir)
```

# 8. Visualization: Multi-Module Heatmap

```{r plot-heatmap}
for (src in names(gene_sets)) {
  # Combine results for this source across all modules
  combined_df <- bind_rows(lapply(names(all_gsea_results), function(m) {
    all_gsea_results[[m]] %>%
      filter(Source == src, padj < 0.05) %>%
      mutate(Module = m) %>%
      select(pathway, NES, Module)
  }))
  
  if (nrow(combined_df) > 1) {
    heatmap_mat <- combined_df %>%
      distinct(pathway, Module, .keep_all = TRUE) %>%
      pivot_wider(names_from = Module, values_from = NES, values_fill = 0) %>%
      column_to_rownames("pathway")

    # Filter pathways present in at least 2 modules
    heatmap_mat <- heatmap_mat[rowSums(heatmap_mat != 0) >= 2, ]
    if(nrow(heatmap_mat) < 2) next
    
    # Auto-scale plot dimensions
    width_in <- max(10, ncol(heatmap_mat) * 0.8)
    height_in <- max(8, nrow(heatmap_mat) * 0.25)
    
    p <- pheatmap(
      as.matrix(heatmap_mat),
      color = colorRampPalette(c("blue", "white", "red"))(100),
      main = paste("Significant Pathway NES Across Modules -", src),
      fontsize_row = 8,
      border_color = "grey60",
      clustering_method = "ward.D2"
    )
    
    ggsave(file.path(plots_dir, paste0("Heatmap_", src, ".pdf")), p, width = width_in, height = height_in, limitsize = FALSE)
    message("Saved heatmap for ", src)
  }
}
```

# 9. Visualization: Dot Plots per Module

```{r plot-dotplots}
plot_fgsea_dot <- function(df, module_name, source_name) {
  sig <- df %>% filter(Source == source_name, padj < 0.05)
  if (nrow(sig) == 0) return(NULL)
  
  top_sig <- sig %>% 
    slice_max(order_by = abs(NES), n = 15) %>%
    mutate(
      clean_pathway = str_wrap(gsub("^(GO_BP_|HALLMARK_|REACTOME_)", "", pathway), width = 50)
    )
    
  ggplot(top_sig, aes(x = NES, y = reorder(clean_pathway, NES), size = -log10(padj), color = NES)) +
    geom_point(alpha = 0.8) +
    scale_color_gradient2(low = "#0571b0", mid = "white", high = "#ca0020") +
    theme_bw(base_size = 10) +
    labs(
      title = paste("Module:", module_name, "| Database:", source_name),
      x = "Normalized Enrichment Score (NES)", y = NULL
    )
}

for (mod in names(all_gsea_results)) {
  for (src in names(gene_sets)) {
    p <- plot_fgsea_dot(all_gsea_results[[mod]], mod, src)
    if (!is.null(p)) {
      ggsave(
        filename = file.path(plots_dir, paste0("DotPlot_", mod, "_", src, ".pdf")),
        plot = p, width = 8, height = 6, useDingbats = FALSE
      )
    }
  }
}
```

# 10. Visualization: Cnet Plots

```{r plot-cnet}
plot_module_network <- function(df, module_name, source_name, n_pathways = 5) {
  sig_res <- df %>% 
    filter(Source == source_name, padj < 0.05) %>% 
    slice_max(order_by = abs(NES), n = n_pathways)
  if(nrow(sig_res) == 0) return(NULL)

  gene_list <- sig_res$leadingEdge
  names(gene_list) <- sig_res$pathway
  
  id_to_symbol <- setNames(feature_map$SYMBOL, feature_map$ENTREZID)
  gene_list_symbols <- lapply(gene_list, function(x) {
    unname(id_to_symbol[as.character(x)]) %>% na.omit()
  })

  cnetplot(gene_list_symbols, 
           showCategory = n_pathways,
           #colorEdge = TRUE,
           #color_edge = "category",
           layout = "circle") +     
    labs(title = paste("Module:", module_name, "| DB:", source_name)) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          plot.margin = margin(20,20,20,20))
}

for (mod in names(all_gsea_results)) {
  for (src in names(gene_sets)) {
    p_net <- plot_module_network(all_gsea_results[[mod]], mod, src)
    if (!is.null(p_net)) {
      ggsave(
        filename = file.path(plots_dir, paste0("Network_", mod, "_", src, ".pdf")),
        plot = p_net, width = 10, height = 8, useDingbats = FALSE
      )
    }
  }
}
```

# 11. Visualization: GSEA Running Score Plots

```{r plot-mountain}
score_plots_dir <- file.path(plots_dir, "GSEA_Scores")
dir.create(score_plots_dir, showWarnings = FALSE)

for (mod in names(all_gsea_results)) {
  # Reconstruct the rank vector for the module
  ranks_df <- data.frame(SOMAmer = rownames(datKME), kME = datKME[[paste0("kME_", mod)]]) %>%
    inner_join(feature_map, by = "SOMAmer") %>%
    filter(is.finite(kME)) %>%
    distinct(ENTREZID, .keep_all = TRUE)
  if(nrow(ranks_df) == 0) next
  gene_ranks <- setNames(ranks_df$kME, ranks_df$ENTREZID)

  for (src in names(gene_sets)) {
    current_sets <- gene_sets[[src]]
    top_paths <- all_gsea_results[[mod]] %>%
      filter(Source == src, padj < 0.05) %>%
      slice_min(order_by = padj, n = 2)
    if (nrow(top_paths) == 0) next

    for (i in 1:nrow(top_paths)) {
      path_name <- top_paths$pathway[i]
      stats <- top_paths[i, ]
      
      p <- plotEnrichment(current_sets[[path_name]], gene_ranks) +
        labs(
          title = paste("Module:", mod, "-", src),
          subtitle = str_wrap(path_name, width = 80),
          caption = paste0("NES: ", round(stats$NES, 2), " | padj: ", format.pval(stats$padj, digits = 2))
        ) +
        theme_bw() +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"))

      clean_path_name <- str_trunc(gsub("[^[:alnum:]]", "_", path_name), 40)
      ggsave(
        filename = file.path(score_plots_dir, paste0("Score_", mod, "_", src, "_", i, "_", clean_path_name, ".pdf")),
        plot = p, width = 8, height = 6, useDingbats = FALSE
      )
    }
  }
}
message("GSEA Mountain plots saved to: ", score_plots_dir)
```