---
title: "Data_preprocessing"
author: "Raymond Otoo"
date: "2026-1-26"
output: pdf_document
---


#=====================================================================================
#
# 00- Set working directory
#=====================================================================================
```{r setup, include=FALSE}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
getwd()
```


#=====================================================================================
#
# 1. Read Protein Expression Matrix and Phenotype data
#=====================================================================================
```{r}
#install.packages("openxlsx")
library(openxlsx)

#z-score noirmalized protein expression matrix
ProtoData <- read.xlsx(file.path(paths$data_dir, "all_ADRC_ADNI_proteomics_samples_Zscore_soma_only.xlsx"))

#phenotypic data
PhenoInfo <- read.xlsx(file.path(paths$data_dir, "all_ADRC_ADNI_csfprot_wmhpheno_CLEAN_FINAL_10_24_2025.xlsx"))


#load PCs 1-3
pc_file <- file.path(paths$pca_dir, "WMH_PCA_scores_PC1_PC3.csv")
if (!file.exists(pc_file)) {
  pc_file <- file.path(paths$data_dir, "WMH_PCA_scores_PC1_PC3.csv")
}
pc1_3 <- read.csv(pc_file)
```

```{r}
pc1_3
```

#Flip PC1 sign for interpretability
```{r}
pc1_3$PC1 <- -pc1_3$PC1

pc1_3
```    



```{r}
colnames(pc1_3)
```


```{r}
PhenoInfo <- cbind(PhenoInfo, pc1_3[,2:4])
PhenoInfo
```
```{r}
colnames(PhenoInfo)
```


```{r}
ProtoData
PhenoInfo
```


```{r}
#SOMA data only
ProtoData <- subset(ProtoData, Panel.SOMA ==1)
PhenoInfo <- subset(PhenoInfo, Panel.SOMA ==1)

ProtoData
PhenoInfo
```

#=====================================================================================
#
# 2. Handling missing Values - Missing value Imputation
# remove peptides with too many missvalues (>90% samples)
#=====================================================================================
```{r}
ProtoData[rowMeans(!is.na(ProtoData)) <0.9,]
ProtoData
```

# Overall missing value count
```{r}
sum(is.na(ProtoData))
```


```{r}
library(imputeLCMD)
 
impute_with_metadata <- function(df, 
                                 model.selector = 1, 
                                 method.MAR = "MLE", 
                                 method.MNAR = "QRILC") {
  # 1. Split into metadata and numeric
  metadata <- df[ , !sapply(df, is.numeric), drop = FALSE]
  numeric_data <- df[ , sapply(df, is.numeric), drop = FALSE]

  # 2. Convert to matrix
  numeric_matrix <- as.matrix(numeric_data)
  # 3. Drop all-NA columns
  numeric_matrix <- numeric_matrix[, colSums(is.na(numeric_matrix)) < nrow(numeric_matrix), drop = FALSE]
  # 4. Drop zero-variance columns
  numeric_matrix <- numeric_matrix[, 
                                   apply(numeric_matrix, 2, function(x) length(unique(na.omit(x))) > 1), 
                                   drop = FALSE]
  # 5. Impute missing values
  numeric_imputed <- impute.MAR.MNAR(numeric_matrix,
                                     model.selector = model.selector,
                                     method.MAR = method.MAR,
                                     method.MNAR = method.MNAR)
  

  # 6. Convert back to dataframe
  numeric_imputed <- as.data.frame(numeric_imputed)
  colnames(numeric_imputed) <- colnames(numeric_matrix)
  # 7. Recombine metadata + imputed numeric data
  result <- cbind(metadata, numeric_imputed)
  return(result )
}
 

#For the combined dataset
ProtoData <- impute_with_metadata(ProtoData)
 
# Check result
dim(ProtoData)
head(ProtoData)
```

# Overall missing value count after imputation
```{r}
sum(is.na(ProtoData))
```
#=====================================================================================
#
# 3. Identify and count duplicates for removal (NB: Some samples had 2 records and hence the 1st occurence of samples with this case #were selected)
#=====================================================================================
```{r}
# 
total_duplicates_ProtoData <- sum(duplicated(ProtoData[,1],fromLast =FALSE))
ProtoData <-ProtoData[!duplicated(ProtoData[,1]),]

total_duplicates_pheno <- sum(duplicated(PhenoInfo[,1],fromLast =FALSE))
PhenoInfo <-PhenoInfo[!duplicated(PhenoInfo[,3]),]

total_duplicates_ProtoData
total_duplicates_pheno
ProtoData
PhenoInfo

```


```{r}
rownames(PhenoInfo) <- PhenoInfo$ID
PhenoInfo$ID <- NULL

rownames(ProtoData) <- ProtoData$ID
ProtoData$ID <- NULL

ProtoData
PhenoInfo
```



```{r}
# are the rownames in PhenoInfo matching ProtoData in the same order?
all(rownames(PhenoInfo) %in% rownames(ProtoData))

# are they in the same order?
all(rownames(PhenoInfo) == rownames(ProtoData))
```
```{r}
colnames(PhenoInfo)
```

```{r}
library(dplyr)

#New filter
PhenoInfo_filtered <- PhenoInfo %>% dplyr::select(1,3,13,21:27,67,28:37,4,40,41:45,50,51,53,54,66,67,68,69,70:72)
colnames(PhenoInfo_filtered)

#save the transformed imputed protein expression matrix to new dataframe
ProtoData_filtered <- ProtoData

PhenoInfo_filtered
ProtoData_filtered
```
```{r}
colnames(PhenoInfo_filtered)
```

#=====================================================================================
#                                                 Focus traits/variables of interest
#
# Sex; (2)
# Age_at_draw; (3)
######## Vascular:
# PC1; (36)
# PC2; (37)
# PC3; (38)
#
# SBP,(10) 
# DM2, (15)
# HTNscore,(35) 
# bmi,(18)
# fsrp,(20) 
# d_cmb,(21) 
# lacunar_stroke,(25) 
# stroke;(26) 
# CAA: caa,(24)
# caa_diag (23)
# probcaa (new variable - 0/1) where probable=1, no or possible = 0; (NEW)
#

######## AD: 
# E4 (allele number: 0, 1, 2);(28) 
# E4status where 0 is E4=0, 1 is E4>0; (30) 
# closest_cdr: 0 = 0, >-0.5 = 1; (31)
# cdr_sob;(32) 
# AD_status;(33) 
# cog_status (based on sumbox >2.5 (34)

########
#=====================================================================================
```{r}
library(dplyr)

#New filter
PhenoInfo_filtered_updated <- PhenoInfo_filtered %>% dplyr::select(1,2,3,36,37,38,10,15,35,18,20,21,25,26,23,24,28,31,32,33,34)
colnames(PhenoInfo_filtered_updated)

PhenoInfo_filtered_updated
```


#binarize Age
```{r}
PhenoInfo_filtered_updated$Sex <- ifelse(PhenoInfo_filtered_updated$Sex =="Male",1,0)
PhenoInfo_filtered_updated
```

#binarize Cohort
```{r}
PhenoInfo_filtered_updated$Cohort <- ifelse(PhenoInfo_filtered_updated$Cohort =="ADNI",1,0)

#change the column name from Cohort to Cohort_bin
colnames(PhenoInfo_filtered_updated)[colnames(PhenoInfo_filtered_updated)=="Cohort"]<-"Cohort_bin"

PhenoInfo_filtered_updated
```


#binarize caa_diag
```{r}
PhenoInfo_filtered_updated$caa_diag <- ifelse(PhenoInfo_filtered_updated$caa_diag =="probable",1,0)

#change the column name from caa_diag to cia_prob
colnames(PhenoInfo_filtered_updated)[colnames(PhenoInfo_filtered_updated)=="caa_diag"]<-"caa_prob"

PhenoInfo_filtered_updated
```

#binarize closest cdr 0 = 0 and > 0= 1
```{r}
PhenoInfo_filtered_updated$closest_cdr <- ifelse(PhenoInfo_filtered_updated$closest_cdr =="0",0,1)
PhenoInfo_filtered_updated
```

#binarizelacunar stroke 0=0 (no) and > 0=1(yes) 
```{r}
PhenoInfo_filtered_updated$lacunar_stroke <- ifelse(PhenoInfo_filtered_updated$lacunar_stroke =="0",0,1)
PhenoInfo_filtered_updated
```
#binarize E4status 
## E4 (allele number: 0, 1, 2);(28) 
# E4status where 0 is E4=0, 1 is E4>0; (30) 
```{r}
PhenoInfo_filtered_updated$E4 <- ifelse(PhenoInfo_filtered_updated$E4 =="0",0,1)

#change the column name from caa_diag to cia_prob
colnames(PhenoInfo_filtered_updated)[colnames(PhenoInfo_filtered_updated)=="E4"]<-"E4status"

PhenoInfo_filtered_updated
```


#=====================================================================================
#
# 4. check the Distinct Levels of PhenoData
#=====================================================================================
```{r}
DistinctLevels <- data.frame(
  Column = names(PhenoInfo_filtered_updated ),
  Levels = sapply(PhenoInfo_filtered_updated , function(x) paste(unique(x), collapse = ", "))
)

DistinctLevels
```


#=====================================================================================
#           Rename phenodata
# 
#=====================================================================================
```{r}
# Make a copy so original remains intact
PhenoInfo_numeric <- PhenoInfo_filtered_updated
print(head(PhenoInfo_numeric))
```

```{r}
library(openxlsx)
 
# Restore original ID and Cohort before saving
PhenoInfo_numeric$Cohort <- PhenoInfo_filtered$Cohort
PhenoInfo_numeric 
```


#save the imputed protein expression matrix and cleaned/reorganized and filtered phenotype data
```{r}
# For multiple objects: 
save(ProtoData, PhenoInfo_numeric , file = file.path(paths$results_dir, "00_output.RData"))

# Save output

write.xlsx(PhenoInfo_numeric, file.path(paths$results_dir, "PhenoInfo_numeric.xlsx"), rowNames = TRUE)
write.xlsx(ProtoData_filtered, file.path(paths$results_dir, "ProtoData_filtered.xlsx"), rowNames = TRUE)

```
