---
title: "new"
author: "Raymond Otoo"
date: "2026-01-30"
output: pdf_document
---


#=====================================================================================
#
#                                           Part A. WGCNA- ADNI_ADRC Combined
# Reference tutorial : https://karobben.github.io/2020/06/10/R/WGCNA_2/
#=====================================================================================

#=====================================================================================
#
# 00- Set working directory
#=====================================================================================
```{r setup, include=FALSE}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
wgcna_dir <- paths$wgcna_dir
getwd()
```


#=====================================================================================
#
# 1. Read Protein Expression Matrix and Phenotype data
#=====================================================================================
```{r}
#Load key results from Part A of ADNI Proteomics analysis
load(file.path(paths$results_dir, "02_output.RData"))
```


```{r}
pheno_combat
Proto_combat_z
```

```{r}
library(WGCNA)
library(dynamicTreeCut)

options(stringsAsFactors = FALSE)
allowWGCNAThreads()
```

```{r}
# Expression matrix: samples x proteins
expr <- Proto_combat_z   # ComBat-corrected, NOT Z-scaled
 
# Phenotype data
traits <- pheno_combat
 
# Ensure matching sample order
expr <- expr[rownames(traits), ]
```


#QC
```{r}
gsg <- goodSamplesGenes(expr, verbose = 3)
expr <- expr[gsg$goodSamples, gsg$goodGenes]
traits <- traits[gsg$goodSamples, ]
```


#samole clustering and outlier detection
```{r}
sampleTree <- hclust(dist(expr), method = "average")
plot(sampleTree, main = "Sample clustering", sub = "", xlab = "")
```

#covariate adjustment
```{r}
#no covariate adjustment
```


#soft-threshold selection
```{r}
powers <- c(1:20)
sft <- pickSoftThreshold(expr, powerVector = powers, verbose = 5)
```

#scale-free topology plot
```{r}
png(file.path(wgcna_dir, "SoftThreshold_ScaleFreeTopology.png"), width = 2000, height = 1500, res = 300)

plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit (signed R²)",
     type = "n",
     main = "Scale independence")
 
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels = powers, col = "red")
abline(h = 0.85, col = "blue", lty = 2)
dev.off()
```


#mean connectivity plot
```{r}
png(file.path(wgcna_dir, "SoftThreshold_MeanConnectivity.png"), width = 2000, height = 1500, res = 300)
 
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = "Mean connectivity")
 
text(sft$fitIndices[,1], sft$fitIndices[,5],
     labels = powers, col = "red")
dev.off()
```
#select power
```{r}
softPower <- 3
```


#Network construction
```{r}
adjacency <- adjacency(
  expr,
  power = softPower,
  type = "signed hybrid"
)
 
TOM <- TOMsimilarity(adjacency, TOMType = "signed")
dissTOM <- 1 - TOM
```

#gene clustering anf module detection
```{r}
geneTree <- hclust(as.dist(dissTOM), method = "average")
```

#gene dendrogram
```{r}
png(file.path(wgcna_dir, "Gene_dendrogram.png"), width = 2000, height = 1500, res = 300)
 
plot(geneTree, main = "Protein clustering on TOM-based dissimilarity",
     xlab = "", sub = "")
dev.off()
```


#6b dynamic tree cut
```{r}
dynamicMods <- cutreeDynamic(
  dendro = geneTree,
  distM = dissTOM,
  deepSplit = 3,
  pamRespectsDendro = FALSE,
  minClusterSize = 30
)
dynamicColors <- labels2colors(dynamicMods)
```


#dendogram + module colors
```{r}
png(file.path(wgcna_dir, "GeneDendrogram_DynamicModules.png"), width = 2000, height = 1500, res = 300)
 
plotDendroAndColors(
  geneTree,
  dynamicColors,
  "Dynamic Tree Cut",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05
)
dev.off()
```
#7. module eigengenes & merging
```{r}
MElist <- moduleEigengenes(expr, colors = dynamicColors)
MEs <- MElist$eigengenes
 
MEDiss <- 1 - cor(MEs)
METree <- hclust(as.dist(MEDiss), method = "average")
```

#7a. Eigengene clustering
```{r}
png("Eigengene_clustering.png", width = 2000, height = 1500, res = 300)
 
plot(METree, main = "Clustering of module eigengenes")
abline(h = 0.30, col = "red")
dev.off()
```

#7b. Merged modules
```{r}
merge <- mergeCloseModules(
  expr,
  dynamicColors,
  cutHeight = 0.30
)
 
moduleColors <- merge$colors
MEs <- merge$newMEs
```

#7c original vs merged modules
```{r}
png("Original_vs_Merged_Modules.png", width = 2000, height = 1500, res = 300)
 
plotDendroAndColors(
  geneTree,
  cbind(dynamicColors, moduleColors),
  c("Original", "Merged"),
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05
)
dev.off()
```


#8. Module-trait relationship(with module labels)
```{r}
traits_interest <- traits[, c(
  "Sex",
  "Age_at_draw",
  "PC1",
  "PC2",
  "PC3",
  "SBP",
  "d_cmb",
  "fsrp",
  "htnscore",
  "lacunar_stroke",
  "dm",
  "bmi",
  "stroke",
  "caa_prob",
  "caa",
  "AD_status",
  "cog_status",
  "E4status"
)]
```


#correlation
```{r}
library(openxlsx)
moduleTraitCor <- cor(MEs, traits_interest, use = "pairwise.complete.obs")
moduleTraitP <- corPvalueStudent(moduleTraitCor, nrow(expr))

#moduleTraitCor
#moduleTraitP

combined_sheet_module_Cor_P <- list("Cor"=moduleTraitCor, "Pval"=moduleTraitP)

# Save output
write.xlsx(combined_sheet_module_Cor_P, file.path(wgcna_dir, "moduleTraitCor_Pval.xlsx"), rowNames = TRUE)
```


#8a. Heatmap
```{r}
textMatrix <- paste0(
  signif(moduleTraitCor, 2), "\n(",
  signif(moduleTraitP, 1), ")"
)
 
dim(textMatrix) <- dim(moduleTraitCor)


png(file.path(wgcna_dir, "Module_Trait_Heatmap.png"), width = 8000, height = 7000, res = 600)
par(mar=c(8,14,4,2))#bottom ,left, top, right 
labeledHeatmap(
  Matrix = moduleTraitCor,
  xLabels = colnames(traits_interest),
  yLabels = names(MEs),
  ySymbols = names(MEs),        # module labels restored
  colorLabels = FALSE,
  colors = blueWhiteRed(50),
  textMatrix = textMatrix,
  setStdMargins = FALSE,
  cex.text = 0.6,
  zlim = c(-1,1),
  main = "Module–Trait Relationships"

)
dev.off()
```
#. Hub protein analysis
```{r}
module <- "blue"   # example
modGenes <- moduleColors == module
 
MM <- cor(expr, MEs[, paste0("ME", module)], use = "p")
GS <- cor(expr, traits_interest$PC1, use = "p")
 
hubCandidates <- which(
  abs(MM) > 0.7 & abs(GS) > 0.3
)
```


## ---------------------------------------------------------
## 9. Save WGCNA_2 outputs for downstream biology + ML
## ---------------------------------------------------------
 
```{r}
save(
  expr,
  traits,
  traits_interest,
  moduleColors,
  MEs,
  moduleTraitCor,
  moduleTraitP,
  file = file.path(wgcna_dir, "WGCNA2_core_outputs.RData")
)
```




```{r}
colnames(MEs)
```
```{r}
MEs
```



```{r}
table(moduleColors)
```
# For multiple objects:
```{r}
save(traits,moduleColors,MEs,expr, file = file.path(wgcna_dir, "03_output.RData"))
```

```{r}
#install.packages("report")
library(report)
report::report_packages()
```

