---
title: "04_Functional_Analysis"
author: "Raymond Otoo"
date: "2026-02-04"
output: pdf_document
---
#=====================================================================================
#
#                                           Part B. Functional Analysis - GO Analysis
#
#=====================================================================================

```{r}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
functional_dir <- paths$functional_dir
dir.create(file.path(functional_dir, "GO_plots"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(functional_dir, "fgsea"), recursive = TRUE, showWarnings = FALSE)
run_go_plots <- FALSE
getwd()
```


#Load libraries & WGCNA outputs
```{r}
#BiocManager::install(c("clusterProfiler", "org.Hs.eg.db"))
library(WGCNA)
library(dplyr)
library(readr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ReactomePA)
library(ggplot2)
library(openxlsx)

#Load key results from Part A of ADNI Proteomics analysis
load(file.path(paths$wgcna_dir, "WGCNA2_core_outputs.RData"))

# expr: samples x proteins
# moduleColors: named vector (proteins)
stopifnot(colnames(expr) == names(moduleColors))


```

#Load SOMAmer annotation + background
```{r}
# Analyte annotation
analyte_info <- read.delim(file.path(paths$data_dir, "CSF_SOMAscan7k_analyte_information.tsv"),sep = "\t", header = TRUE, check.names = FALSE)

# Background universe
background <- read.csv(file.path(paths$data_dir, "background_soma.csv"))
background_soma <- background$EntrezGeneSymbol

length(background_soma)
head(analyte_info)
```

#Map SOMAmer → SYMBOL → ENTREZ
```{r}
all_features <- colnames(expr)
 
feature_key <- analyte_info %>%
  dplyr::filter(Analytes %in% all_features) %>%
  dplyr::select(Analytes, EntrezGeneSymbol) %>%
  dplyr::distinct() %>%
  dplyr::rename(
    SOMAmer = Analytes,
    SYMBOL  = EntrezGeneSymbol
  )
 
map_all <- bitr(
  feature_key$SYMBOL,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
) %>%
  dplyr::distinct(SYMBOL, .keep_all = TRUE)
 
universe_entrez <- map_all$ENTREZID
```

#Module → Gene tables (CSV + Excel)
```{r}
modules_to_export <- setdiff(unique(moduleColors), "grey")

module_gene_list <- list()

for (mod in modules_to_export) {

  somamers <- colnames(expr)[moduleColors == mod]

module_df <- feature_key %>%
  dplyr::filter(SOMAmer %in% somamers) %>%
  dplyr::left_join(map_all, by = "SYMBOL") %>%
  dplyr::mutate(Module = mod) %>%
  dplyr::select(Module, SOMAmer, SYMBOL, ENTREZID)

  module_gene_list[[mod]] <- module_df
}

final_gene_list_df <- bind_rows(module_gene_list)
write_csv(final_gene_list_df, file.path(functional_dir, "Module_Gene_Lists.csv"))
```

#Column-wise Excel
```{r}
module_gene_vectors <- lapply(modules_to_export, function(mod) {

  somamers <- colnames(expr)[moduleColors == mod]

  feature_key %>%
    filter(SOMAmer %in% somamers) %>%
    inner_join(map_all, by = "SYMBOL") %>%
    pull(SYMBOL) %>%
    unique()
})

names(module_gene_vectors) <- modules_to_export

max_len <- max(lengths(module_gene_vectors))
padded <- lapply(module_gene_vectors, function(x) c(x, rep(NA, max_len - length(x))))
final_transposed_df <- as.data.frame(padded)

write.xlsx(
  final_transposed_df,
  file.path(functional_dir, "Module_Genes_Columnwise.xlsx"),
  overwrite = TRUE
)

```

#Functional enrichment (GO-BP only in this notebook)
```{r}
plot_go_for_module <- function(module_color) {

  somamers <- colnames(expr)[moduleColors == module_color]

  entrez <- feature_key %>%
    filter(SOMAmer %in% somamers) %>%
    inner_join(map_all, by = "SYMBOL") %>%
    pull(ENTREZID) %>%
    unique()

  if (length(entrez) == 0) return(NULL)

  ego <- enrichGO(
    gene = entrez,
    universe = universe_entrez,
    OrgDb = org.Hs.eg.db,
    keyType = "ENTREZID",
    ont = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff = 0.1,
    readable = TRUE
  )

  if (is.null(ego) || nrow(ego@result) == 0) return(NULL)

  go_df <- ego@result %>%
    filter(p.adjust <= 0.1) %>%
    arrange(p.adjust) %>%
    head(10)

  if (nrow(go_df) == 0) return(NULL)

  ggplot(go_df, aes(GeneRatio, Description)) +
    geom_point(aes(size = Count, color = p.adjust)) +
    scale_color_gradient(low = "red", high = "blue") +
    theme_bw() +
    labs(
      title = paste("GO-BP Enrichment – Module", module_color),
      x = "Gene Ratio",
      y = NULL
    )
}
```

#Run enrichment for all modules
```{r}
if (run_go_plots) {
  dir.create(functional_dir, showWarnings = FALSE)
  dir.create(file.path(functional_dir, "GO_plots"), showWarnings = FALSE)

  for (mod in modules_to_export) {
    p <- plot_go_for_module(mod)
    if (!is.null(p)) {
      ggsave(
        file.path(functional_dir, "GO_plots", paste0("GO_", mod, ".png")),
        p, width = 10, height = 6
      )
    }
  }
} else {
  message("Skipping GO plot generation in 04_Functional_Analysis.Rmd. Use 07a/07b for enrichment workflows.")
}

```

```{r}
# The original mixed notebook continued into an embedded fgsea workflow.
# That logic is now executed in the dedicated original notebooks:
# 07a_Functional_Analysis_GSEA.Rmd and 07b_Funtional_Analysis_ORA.Rmd.
knitr::knit_exit()
```


#=====================================================================================
#
#                                           Part B. GSEA - Fgsea
#
#=====================================================================================
```{r}
library(fgsea)
library(msigdbr)
```



```{r}
ls()
```
```{r}
#-----------------------------------------
# Recompute module membership (kME)
#-----------------------------------------
 
# expr: samples x proteins
# MEs : samples x module eigengenes
 
stopifnot(all(colnames(expr) == names(moduleColors)))
 
datKME <- as.data.frame(
  cor(expr, MEs, use = "p")
)
 
colnames(datKME) <- paste0("kME", substring(colnames(MEs), 3))
rownames(datKME) <- colnames(expr)
```

```{r}
datKME
```




#-----------------------------------------
# Create ranked gene lists for fgsea
#-----------------------------------------
 
# datKME assumed present from WGCNA outputs
# rows: proteins (SOMAmers)
# columns: kME values per module
```{r}
stopifnot(rownames(datKME) %in% feature_key$SOMAmer)
```


#-----------------------------------------
# Load GO gene sets (fgsea-compatible)
#-----------------------------------------
```{r}
go_bp <- msigdbr(
  species = "Homo sapiens",
  category = "C5",
  subcategory = "BP"
)
 
go_pathways <- split(go_bp$entrez_gene, go_bp$gs_name)
```

#-----------------------------------------
# fgsea per module using kME ranking
#-----------------------------------------
```{r}
run_fgsea_for_module <- function(module_color) {
 
  kme_col <- paste0("kME", module_color)
 
  if (!kme_col %in% colnames(datKME)) return(NULL)
 
  #--------------------------------------------------

  # Build ranked SOMAmer list

  #--------------------------------------------------

  rank_df <- data.frame(

    SOMAmer = rownames(datKME),

    kME     = datKME[, kme_col]

  ) %>%

    filter(!is.na(kME)) %>%

    arrange(desc(kME))
 
  #--------------------------------------------------

  # Map SOMAmer → SYMBOL → ENTREZ

  #--------------------------------------------------

  rank_df <- rank_df %>%

    left_join(feature_key, by = "SOMAmer") %>%

    left_join(map_all, by = "SYMBOL") %>%

    filter(!is.na(ENTREZID)) %>%

    distinct(ENTREZID, .keep_all = TRUE)
 
  if (nrow(rank_df) < 50) return(NULL)
 
  #--------------------------------------------------

  # Create ranked vector

  #--------------------------------------------------

  gene_ranks <- rank_df$kME

  names(gene_ranks) <- rank_df$ENTREZID

  gene_ranks <- sort(gene_ranks, decreasing = TRUE)
 
  #--------------------------------------------------

  # Run fgsea

  #--------------------------------------------------
  
  fg <- fgseaMultilevel(
  pathways     = go_sets,
  stats        = gene_ranks,
  minSize      = 15,
  maxSize      = 500,
  nPermSimple  = 10000,   # handles unbalanced pathways
  eps          = 0,
  nproc        = 1        # 🔥 prevents serialize() crashes
)
  
  fg

}

 
```


```{r}
library(msigdbr)
 
msig_go <- msigdbr(
  species = "Homo sapiens",
  category = "C5",
  subcategory = "BP"
)
 
go_sets <- split(msig_go$entrez_gene, msig_go$gs_name)
 
# Restrict to universe
go_sets <- lapply(go_sets, intersect, universe_entrez)
go_sets <- go_sets[sapply(go_sets, length) >= 15]
```








#-----------------------------------------
# Run fgsea across modules
#-----------------------------------------
```{r}
fgsea_results <- list()
 
for (mod in modules_to_export) {
 
  fg <- run_fgsea_for_module(mod)
 
  if (!is.null(fg)) {
    fgsea_results[[mod]] <- fg %>%
      arrange(padj) %>%
      filter(padj < 0.05)
  }
}
```

#-----------------------------------------
# Save fgsea results
#-----------------------------------------
```{r}
 fgsea_dir <- file.path(functional_dir, "fgsea")
 dir.create(fgsea_dir, showWarnings = FALSE)
 
for (mod in names(fgsea_results)) {
 
  write_csv(
    fgsea_results[[mod]],
    file.path(fgsea_dir, paste0("fgsea_GO_BP_", mod, ".csv"))
  )
}
```


```{r}
plot_fgsea_for_module <- function(fgsea_df, module_color) {
 
  top_terms <- fgsea_df %>% head(15)
 
  ggplot(top_terms,
         aes(x = NES, y = reorder(pathway, NES))) +
    geom_point(aes(size = -log10(padj), color = NES)) +
    theme_bw() +
    labs(
      title = paste("fgsea GO-BP – Module", module_color),
      x = "Normalized Enrichment Score",
      y = "GO Biological Process"
    )
}
```


```{r}
for (mod in names(fgsea_results)) {
 
  p <- plot_fgsea_for_module(fgsea_results[[mod]], mod)
 
  ggsave(
    file.path(fgsea_dir, paste0("fgsea_", mod, ".png")),
    p, width = 10, height = 7
  )
}
```


#-----------------------------------------
#               FGSEA - Using Hallmark
#-----------------------------------------
```{r}
library(dplyr)
library(fgsea)
library(msigdbr)
library(org.Hs.eg.db)
library(clusterProfiler)
```

```{r}
all_features <- colnames(expr)
 
feature_key <- analyte_info %>%
  dplyr::filter(Analytes %in% all_features) %>%
  dplyr::select(Analytes, EntrezGeneSymbol) %>%
  dplyr::distinct() %>%
  dplyr::rename(
    SOMAmer = Analytes,
    SYMBOL  = EntrezGeneSymbol
  )
 
map_all <- bitr(
  feature_key$SYMBOL,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
) %>%
  dplyr::distinct(SYMBOL, .keep_all = TRUE)
 
universe_entrez <- map_all$ENTREZID

feature_map <- feature_key %>%
  inner_join(map_all, by = "SYMBOL")
```

```{r}
stopifnot(all(colnames(expr) == names(moduleColors)))
 
datKME <- as.data.frame(
  cor(expr, MEs, use = "p")
)
 
colnames(datKME) <- paste0("kME_", sub("^ME", "", colnames(MEs)))
rownames(datKME) <- colnames(expr)
```

```{r}
hallmark_df <- msigdbr(
  species = "Homo sapiens",
  category = "H"
)
 
hallmark_sets <- hallmark_df %>%
  split(x = .$entrez_gene, f = .$gs_name)
```


```{r}
reactome_df <- msigdbr(
  species = "Homo sapiens",
  category = "C2",
  subcategory = "CP:REACTOME"
)
 
reactome_sets <- reactome_df %>%
  split(x = .$entrez_gene, f = .$gs_name)
```


```{r}
reactome_df <- msigdbr(

  species = "Homo sapiens",
  category = "C2",
  subcategory = "CP:REACTOME"
)
 
reactome_sets <- reactome_df %>%
  split(x = .$entrez_gene, f = .$gs_name)

 
run_fgsea_for_module <- function(module,
                                 pathway_sets,
                                 minSize = 15,
                                 maxSize = 500) {
 
  kme_col <- paste0("kME_", module)
 
  if (!kme_col %in% colnames(datKME)) {
    message("Skipping module ", module, ": kME not found")
    return(NULL)
  }
 
  rank_df <- data.frame(
    SOMAmer = rownames(datKME),
    kME = datKME[, kme_col]
  ) %>%

    inner_join(feature_map, by = "SOMAmer") %>%
    filter(!is.na(kME), !is.na(ENTREZID)) %>%
    distinct(ENTREZID, .keep_all = TRUE)
 
  gene_ranks <- rank_df$kME
  names(gene_ranks) <- rank_df$ENTREZID
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)
 
  fg <- fgseaMultilevel(
  pathways     = go_sets,
  stats        = gene_ranks,
  minSize      = 15,
  maxSize      = 500,
  nPermSimple  = 10000,   # handles unbalanced pathways
  eps          = 0,
  nproc        = 1        # prevents serialize() crashes
)
}
 
```



```{r}
fgsea_results <- list()
 
for (mod in modules_to_export) {
  fg <- run_fgsea_for_module(
    module = mod,
    pathway_sets = hallmark_sets
  )
 
  if (!is.null(fg)) {
    fgsea_results[[mod]] <- fg %>%
      arrange(padj)
  }
}
```
```{r}
fgsea_results <- list()
 
for (mod in modules_to_export) {
  fg <- run_fgsea_for_module(
    module = mod,
    pathway_sets = reactome_sets
  )
 
  if (!is.null(fg)) {
    fgsea_results[[mod]] <- fg %>%
      arrange(padj)
  }
}
```

```{r}
plot_fgsea_dot <- function(fg, module_name) {
 
  fg %>%
    filter(padj < 0.05) %>%
    ggplot(aes(
      x = NES,
      y = reorder(pathway, NES),
      size = -log10(padj),
      color = NES
    )) +
    geom_point() +
    scale_color_gradient2(
      low = "blue", mid = "white", high = "red"
    ) +
    theme_bw() +
    labs(
      title = paste("fgsea – Module", module_name),
      y = "Pathway",
      x = "Normalized Enrichment Score (NES)"
    )
}
```

```{r}
plot_fgsea_dot(fgsea_results[["blue"]], "blue")
```


```{r}
prep_fgsea <- function(fg, source) {
  fg %>%
    filter(padj < 0.05) %>%
    select(pathway, NES) %>%
    mutate(Source = source)
}
```

```{r}
# Define the helper function if you haven't already
prep_fgsea <- function(fg_res, source_label) {
  if (is.null(fg_res) || nrow(fg_res) == 0) return(NULL)
  fg_res %>%
    as_tibble() %>%
    filter(padj < 0.05) %>%
    dplyr::select(pathway, NES, padj) %>%
    mutate(Source = source_label)
}
 
# List of your result objects and their corresponding labels
result_lists <- list(
  "Hallmark" = hallmark_sets,
  "Reactome" = reactome_sets,
  "GO"       = go_sets
)
 
# Combine for the "blue" module automatically
target_module <- "blue"
 
combined_fgsea <- lapply(names(result_lists), function(label) {
  res_list <- result_lists[[label]]
  prep_fgsea(res_list[[target_module]], label)
}) %>% 
  bind_rows()
```



```{r}
colnames(result_lists[[1]])
```
```{r}
names(result_lists)
summary(result_lists[[1]]$padj)
str(result_lists, max.level = 2)
```




```{r}
# Initialize a structured results list
# Structure: gsea_db_results$Database$Module
gsea_db_results <- list(
  "Hallmark" = list(),
  "Reactome" = list(),
  "GO_BP"    = list()
)
 
# Run GSEA for each database across all modules
for (mod in modules_to_export) {
  # Run Hallmark
  gsea_db_results$Hallmark[[mod]] <- run_fgsea_for_module(mod, hallmark_sets)
  # Run Reactome
  gsea_db_results$Reactome[[mod]] <- run_fgsea_for_module(mod, reactome_sets)
  # Run GO_BP
  gsea_db_results$GO_BP[[mod]] <- run_fgsea_for_module(mod, go_sets)
}
```


```{r}
save_gsea_by_database <- function(results_object, 
                                  outdir = file.path(functional_dir, "fgsea"), 
                                  padj_cutoff = 0.05) {
  if (!dir.exists(outdir)) {
    dir.create(outdir, recursive = TRUE)
  }
 
  for (src in names(results_object)) {
    wb <- createWorkbook()
    cat("Processing Database:", src, "\n")
    module_data <- results_object[[src]]
    # Check if this database has any modules at all
    if (length(module_data) == 0) {
      cat("No results found for", src, "- skipping.\n")
      next
    }
 
    for (mod in names(module_data)) {
      df <- module_data[[mod]]
      # Skip if NULL or empty
      if (is.null(df) || nrow(df) == 0) next
      # Filter by significance
      df <- df %>% filter(padj < padj_cutoff)
      if (nrow(df) > 0) {
        # Convert leadingEdge list-column to string so it can be saved in Excel
        if ("leadingEdge" %in% colnames(df)) {
          df$leadingEdge <- sapply(df$leadingEdge, paste, collapse = ", ")
        }
        # Add worksheet and write data
        addWorksheet(wb, mod)
        writeData(wb, mod, df)
      }
    }
    # Only save if the workbook actually has sheets
    if (length(names(wb)) > 0) {
      file_path <- file.path(outdir, paste0("ORA_", src, "_Results.xlsx"))
      saveWorkbook(wb, file = file_path, overwrite = TRUE)
      cat("Saved:", file_path, "\n")
    } else {
      cat("No significant results for", src, "to save.\n")
    }
  }
  cat("Finished exporting results.\n")
}
```



```{r}
# Call the function with the results object
save_gsea_by_database(gsea_db_results)
```


#==============================================================
# Save GSEA Results Separately Per Database
#==============================================================
```{r}
library(openxlsx)
library(dplyr)
 
save_gsea_by_database <- function(result_lists,
                                  outdir = file.path(functional_dir, "fgsea"),
                                  padj_cutoff = 0.05) {
  if (!dir.exists(outdir)) {
    dir.create(outdir, recursive = TRUE)
  }
  # Loop over databases directly
  for (src in names(result_lists)) {
    wb <- createWorkbook()
    cat("Creating workbook for:", src, "\n")
    module_list <- result_lists[[src]]
    for (mod in names(module_list)) {
      df <- module_list[[mod]]
      # Skip if NULL
      if (is.null(df)) next
      # Make sure it's a dataframe
      df <- as.data.frame(df)
      # Detect correct p-value column
      pcol <- if ("padj" %in% colnames(df)) {
        "padj"
      } else if ("p.adjust" %in% colnames(df)) {
        "p.adjust"
      } else {
        next
      }
      df <- df[df[[pcol]] < padj_cutoff, ]
      if (nrow(df) > 0) {
        addWorksheet(wb, mod)
        writeData(wb, mod, df)
      }
    }
    saveWorkbook(
      wb,
      file = paste0(outdir, "ORA_", src, "_Results.xlsx"),
      overwrite = TRUE
    )
  }
  cat("Finished exporting database-specific GSEA results.\n")
}

```

```{r}
save_gsea_by_database(result_lists)
```

