---
title: "05_Network_Analysis"
author: "Raymond Otoo"
date: "2026-02-03"
output: pdf_document
---

#=====================================================================================
#
# 00- Set working directory
#=====================================================================================
```{r setup, include=FALSE}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
getwd()
```


#=====================================================================================
#
# 1. Read Protein Expression Matrix and Phenotype data
#=====================================================================================
```{r}
#Load key results from Part A of ADNI Proteomics analysis
load(file.path(paths$wgcna_dir, "WGCNA2_core_outputs.RData"))
```


############################################################
## WGCNA module–trait annotation and protein signature derivation
## no cohort residualization
## Rationale: cohort residualization had minimal impact on structure
############################################################
```{r}
## Libraries
library(dplyr)
library(tibble)
```

 
## ---------------------------------------------------------
## 1. Read module–trait correlation and p-value matrices
## ---------------------------------------------------------
```{r} 
cor_mat <- moduleTraitCor
pval_mat <- moduleTraitP
``` 


## ---------------------------------------------------------
## 2. Define biologically motivated trait groups
## ---------------------------------------------------------
```{r} 
# Vascular / CSVD (PC1 = strongest WMH proteomic signal)
vascular_traits <- c(
  "PC1", 
  "fsrp", 
  "htnscore", 
  "SBP", 
  "d_cmb", 
  "lacunar_stroke"
)
 
# Cerebral amyloid angiopathy
caa_traits <- c(
  "caa_prob", 
  "caa")
 
# AD / amyloid / cognition
ad_traits <- c(
  "AD_status", 
  "E4status", 
  "cog_status")

``` 

```{r}
summary(abs(cor_mat[, vascular_traits]))
summary(abs(cor_mat[, caa_traits]))
summary(abs(cor_mat[, ad_traits]))


#summary(pval_mat[, vascular_traits])
```
```{r}
hist(abs(cor_mat[, "PC1"]), breaks = 30, main = "Module–PC1 correlations")
```



## ---------------------------------------------------------
## 3. Multiple testing correction (FDR per trait)
## ---------------------------------------------------------
```{r} 
qval_mat <- pval_mat

for (trait in colnames(pval_mat)) {
  qval_mat[, trait] <- p.adjust(pval_mat[, trait], method = "BH")
}
 
## Thresholds for defining meaningful module–trait associations
r_thresh <- 0.12
q_thresh <- 0.05
```

 
## ---------------------------------------------------------
## 4. Helper function: count significant trait "hits" per module
##    (sign-aware)
## ---------------------------------------------------------
```{r} 
score_modules <- function(trait_set,
                          direction = c("two_sided", "positive", "negative")) {
  direction <- match.arg(direction)
  sapply(rownames(cor_mat), function(mod) {
 
    r_vec <- as.numeric(cor_mat[mod, trait_set])
    q_vec <- as.numeric(qval_mat[mod, trait_set])
 
    if (direction == "two_sided") {
      sum(abs(r_vec) >= r_thresh & q_vec < q_thresh, na.rm = TRUE)

    } else if (direction == "positive") {
      sum(r_vec >= r_thresh & q_vec < q_thresh, na.rm = TRUE)

    } else {
      sum(r_vec <= -r_thresh & q_vec < q_thresh, na.rm = TRUE)
    }
  })
}
``` 


## ---------------------------------------------------------
## 5. Compute biology-specific module hit counts
## ---------------------------------------------------------
```{r} 
vascular_hits <- score_modules(vascular_traits, direction = "positive")
caa_hits      <- score_modules(caa_traits,      direction = "positive")
ad_hits_pos   <- score_modules(ad_traits,       direction = "positive")
ad_hits_neg   <- score_modules(ad_traits,       direction = "negative")
module_scores <- data.frame(
  module        = rownames(cor_mat),
  vascular_hits = vascular_hits,
  caa_hits      = caa_hits,
  ad_hits_pos   = ad_hits_pos,
  ad_hits_neg   = ad_hits_neg,
  stringsAsFactors = FALSE
)
``` 


## ---------------------------------------------------------
## 6. Manual biological labeling (sign-aware interpretation)
## ---------------------------------------------------------
```{r} 
module_scores$label_auto <- "none"

# Core vascular / CSVD–hypertension modules 
module_scores$label_auto[
  module_scores$vascular_hits >= 2
] <- "vascular"

# Mixed vascular + AD + CAA biology 
module_scores$label_auto[
  module_scores$caa_hits >= 1
] <- "CAA"

# AD resilience / healthy aging module 
module_scores$label_auto[
  module_scores$ad_hits_neg >= 1
] <- "protective_healthy"
``` 


```{r}
module_scores
```

```{r}
#---------------------------------------------------------
# Save module scoring table
#---------------------------------------------------------

dir.create("results/network_annotation",
           recursive = TRUE,
           showWarnings = FALSE)

write.csv(
  module_scores,
  "results/network_annotation/Module_Trait_Scoring_Table.csv",
  row.names = FALSE
)

```




############################################################
## Alternative Continuous Domain Scoring Framework
## (Magnitude-aware, FDR-controlled, no arbitrary r threshold)
############################################################
```{r}


library(dplyr)

#---------------------------------------------------------
# 1. Ensure FDR matrix exists
#---------------------------------------------------------

if (!exists("qval_mat")) {
  qval_mat <- pval_mat
  for (trait in colnames(pval_mat)) {
    qval_mat[, trait] <- p.adjust(pval_mat[, trait], method = "BH")
  }
}

q_thresh <- 0.05

#---------------------------------------------------------
# 2. Continuous scoring function
#    Sum of absolute correlations for FDR-significant traits
#---------------------------------------------------------

score_modules_continuous <- function(trait_set) {
  sapply(rownames(cor_mat), function(mod) {

    r_vec <- as.numeric(cor_mat[mod, trait_set])
    q_vec <- as.numeric(qval_mat[mod, trait_set])

    sum(abs(r_vec[q_vec < q_thresh]), na.rm = TRUE)
  })
}

#---------------------------------------------------------
# 3. Compute continuous domain scores
#---------------------------------------------------------

vascular_score <- score_modules_continuous(vascular_traits)
caa_score      <- score_modules_continuous(caa_traits)
ad_score       <- score_modules_continuous(ad_traits)

module_scores_continuous <- data.frame(
  module         = rownames(cor_mat),
  vascular_score = vascular_score,
  caa_score      = caa_score,
  ad_score       = ad_score,
  stringsAsFactors = FALSE
)

#---------------------------------------------------------
# 4. Standardize scores (z-scores)
#---------------------------------------------------------

module_scores_continuous <- module_scores_continuous %>%
  mutate(
    vascular_z = scale(vascular_score)[,1],
    caa_z      = scale(caa_score)[,1],
    ad_z       = scale(ad_score)[,1]
  )

#---------------------------------------------------------
# 5. Assign dominant biological domain (based on z-score magnitude)
#---------------------------------------------------------

module_scores_continuous$label_auto_continuous <- apply(
  module_scores_continuous[, c("vascular_z","caa_z","ad_z")],
  1,
  function(x) {
    domains <- c("vascular","CAA","AD")
    domains[which.max(abs(x))]
  }
)

#---------------------------------------------------------
# 6. Save results for comparison
#---------------------------------------------------------

dir.create("results/network_annotation",
           recursive = TRUE,
           showWarnings = FALSE)

write.csv(
  module_scores_continuous,
  "results/network_annotation/Module_Trait_Scoring_Continuous.csv",
  row.names = FALSE
)

#---------------------------------------------------------
# 7. Display side-by-side comparison with original scoring
#---------------------------------------------------------

comparison_table <- module_scores %>%
  left_join(module_scores_continuous,
            by = "module")

print(comparison_table)


```

