---
title: "06_ADNI_ADRC_ML"
author: "Raymond Otoo"
date: "2026-01-30"
output: pdf_document
---


#=====================================================================================
#
# 00- Set working directory
#=====================================================================================
```{r setup, include=FALSE}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
ml_dir <- file.path(paths$results_dir, "ml")
dir.create(ml_dir, recursive = TRUE, showWarnings = FALSE)
getwd()
```


#=====================================================================================
#
# 1. Read saved object containing Traits and MEs
#=====================================================================================
```{r}
#Load key results from Part A of ADNI Proteomics analysis
load(file.path(paths$wgcna_dir, "03_output.RData"))
```


#=====================================================================================
#
#                                   2. ML implementation

# 1. Feature Matrix Preparation
#=====================================================================================
```{r}
# Ensure row alignment
stopifnot(rownames(MEs) == rownames(traits))
 
# Combine features + outcome
ml_df <- data.frame(
  MEs,
  stroke = factor(traits$stroke)  # change outcome if needed
)
 
# Remove samples with missing outcome
ml_df <- ml_df[!is.na(ml_df$stroke), ]
```



```{r}
ml_df
```



```{r}

#do once
#install.packages("caret",dependencies = TRUE)
```


## 2. Train/test split(No leakage)
```{r}
set.seed(42)
 
train_idx <- caret::createDataPartition(
  ml_df$stroke,
  p = 0.7,
  list = FALSE
)
 
train_df <- ml_df[train_idx, ]
test_df  <- ml_df[-train_idx, ]
```

```{r}
train_df
test_df
```



# 3. Feature scaling 
```{r}
preproc <- caret::preProcess(
  train_df[, -ncol(train_df)],
  method = c("center", "scale")
)
 
X_train <- predict(preproc, train_df[, -ncol(train_df)])
X_test  <- predict(preproc, test_df[, -ncol(test_df)])
 
y_train <- train_df$stroke
y_test  <- test_df$stroke
```



# 4. Elastic Net(Primary Model)
# * Handles correlated modules
# * Sparse
# * Interpretable
# * Biological interpretable
```{r}
library(glmnet)
 
x_mat <- as.matrix(X_train)
y_vec <- y_train
 
cv_fit <- cv.glmnet(
  x = x_mat,
  y = y_vec,
  family = "binomial",
  alpha = 0.5,          # Elastic Net
  nfolds = 5
)
 
best_lambda <- cv_fit$lambda.min
```

#5. Evaluate on held-out test set
```{r}
# Predict probabilities
test_probs <- predict(
  cv_fit,
  newx = as.matrix(X_test),
  s = best_lambda,
  type = "response"
)
 
# Convert to class labels
test_pred <- ifelse(test_probs > 0.5, "1", "0")
test_pred <- factor(test_pred, levels = levels(y_test))
 
# Confusion matrix
caret::confusionMatrix(test_pred, y_test)
```
#ROC
```{r}
library(pROC)
 
roc_obj <- roc(
  response = y_test,
  predictor = as.numeric(test_probs)
)
 
auc(roc_obj)
plot(roc_obj)
```


# 6. Interpret model
```{r}
coef_df <- as.data.frame(as.matrix(
  coef(cv_fit, s = best_lambda)
))
 
coef_df$Module <- rownames(coef_df)
colnames(coef_df)[1] <- "Coefficient"
 
# Remove intercept
coef_df <- coef_df[coef_df$Module != "(Intercept)", ]
 
# Rank important modules
coef_df <- coef_df[order(abs(coef_df$Coefficient), decreasing = TRUE), ]
coef_df 
head(coef_df)
```

# 7 . Map back the hub proteins( biological grounding)
```{r}
top_module <- gsub("ME", "", coef_df$Module[1])
 
hubGenes <- names(expr)[moduleColors == top_module]
head(hubGenes)

top_module
hubGenes
```
