---s
title: "funct_new"
author: "Raymond Otoo"
date: "2026-02-10"
output: pdf_document
---

#=====================================================================================
#                             Functional Analysis/Network Analysis - Summary of the Workflow
#
# 1.Annotation: Maps SOMAmers to Entrez IDs.
# 2.Ranking: Uses WGCNA kME values to rank every protein in every module.
# 3.Enrichment: Runs fgseaMultilevel against three distinct databases (Hallmark, Reactome, GO).
# 4.Integration: Combines them into a single data structure.
# 5.Visualization: Creates heatmaps to compare modules and faceted dot plots to inspect individual modules.
#=====================================================================================


#==============================================================
# 1. Setup
#==============================================================
```{r}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
gsea_dir <- paths$gsea_dir
gsea_plots_dir <- file.path(gsea_dir, "plots")

suppressPackageStartupMessages({
  library(WGCNA)
  library(dplyr)
  library(readr)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(ReactomePA)
  library(ggplot2)
  library(openxlsx)
  library(fgsea)
  library(msigdbr)
  library(tidyr)
  library(tibble)
  library(pheatmap)
})

dir.create(gsea_plots_dir, recursive = TRUE, showWarnings = FALSE)

```


#==============================================================
# 2. Load WGCNA Results
#==============================================================
```{r}
# Loads previously saved r object froom WGCNA containing results:
# expr, moduleColors, MEs
load(file.path(paths$wgcna_dir, "WGCNA2_core_outputs.RData"))

```


```{r}
# Make sure names are correct
if (is.null(names(moduleColors))) names(moduleColors) <- colnames(expr)
stopifnot(identical(colnames(expr), names(moduleColors)))
 
# List of modules (excluding grey)
modules <- setdiff(unique(moduleColors), "grey")
 
# Compute kME
datKME <- cor(expr, MEs, use = "p")
datKME <- as.data.frame(datKME)
colnames(datKME) <- paste0("kME_", sub("^ME", "", colnames(MEs)))
rownames(datKME) <- colnames(expr)
```


#==============================================================
# 3. Annotation Mapping
#==============================================================
```{r}
# Load feature information
analyte_info <- read.delim(
  file.path(paths$data_dir, "CSF_SOMAscan7k_analyte_information.tsv"),
  sep = "\t",
  header = TRUE,
  check.names = FALSE
)
 
# Keep only overlapping SOMAmers
feature_key <- analyte_info %>%
  dplyr::filter(Analytes %in% colnames(expr)) %>%
  dplyr::select(Analytes, EntrezGeneSymbol) %>%
  dplyr::distinct() %>%
  dplyr::rename(SOMAmer = Analytes, SYMBOL = EntrezGeneSymbol)
 
# Map SYMBOL → ENTREZID
symbol_map <- clusterProfiler::bitr(
  feature_key$SYMBOL,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>% distinct(SYMBOL, .keep_all = TRUE)
 
# Join to get final mapping
feature_map <- inner_join(feature_key, symbol_map, by = "SYMBOL")
 
cat("Number of mapped genes:", nrow(feature_map), "\n")
```

```{r}
nrow(feature_key)
nrow(feature_map)
head(analyte_info$Analytes)
```
```{r}
dim(expr)
dim(MEs)
```

#==============================================================
# 4. Load Gene Sets
#==============================================================
```{r}
library(msigdbr)
 
# 1. Download the gene sets
hm <- msigdbr(species = "Homo sapiens", category = "H")
react <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")
go <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")
 
# 2. Build pathway lists with proper names
hallmark_sets <- split(hm$entrez_gene, as.character(hm$gs_name))
reactome_sets <- split(react$entrez_gene, as.character(react$gs_name))
go_sets <- split(go$entrez_gene, as.character(go$gs_name))
 
# 3. Check
length(hallmark_sets)      # should be ~50 for Hallmark
head(names(hallmark_sets)) # descriptive names
length(reactome_sets)
head(names(reactome_sets))
length(go_sets)
head(names(go_sets))
```


#==============================================================
# 5. GSEA Function
#==============================================================
```{r}
run_fgsea_module <- function(module, pathway_sets) {

  kme_col <- paste0("kME_", module)

  module_genes <- names(moduleColors[moduleColors == module])

  ranks_df <- data.frame(
    SOMAmer = rownames(datKME),
    kME = datKME[, kme_col]
  ) %>%
    filter(SOMAmer %in% module_genes) %>%
    inner_join(feature_map, by = "SOMAmer") %>%
    filter(is.finite(kME)) %>%
    distinct(ENTREZID, .keep_all = TRUE)

  if (nrow(ranks_df) < 15) return(NULL)

  gene_ranks <- ranks_df$kME
  names(gene_ranks) <- ranks_df$ENTREZID
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)

  fgseaMultilevel(
    pathways = pathway_sets,
    stats = gene_ranks,
    minSize = 5,
    maxSize = 500,
    scoreType = "pos",
    nproc = 1 #force single-threaded
  )
}

```


#check how many Module genes map to entrez sets
```{r}
# For each module, count mapped genes
module_gene_counts <- sapply(modules, function(mod) {
  module_genes <- names(moduleColors[moduleColors == mod])
  mapped_genes <- feature_map$ENTREZID[feature_map$SOMAmer %in% module_genes]
  length(unique(mapped_genes))
})
 
module_gene_counts
```

## Run Unit Test for one mudule
```{r}
library(dplyr)
library(fgsea)
 
# Pick a large module, e.g., "green"
mod <- "green"
kme_col <- paste0("kME_", mod)
module_genes <- names(moduleColors[moduleColors == mod])
 
ranks_df <- data.frame(
  SOMAmer = rownames(datKME),
  kME = datKME[, kme_col]
) %>%
  filter(SOMAmer %in% module_genes) %>%
  inner_join(feature_map, by = "SOMAmer") %>%
  filter(is.finite(kME)) %>%
  distinct(ENTREZID, .keep_all = TRUE)
 
gene_ranks <- ranks_df$kME
names(gene_ranks) <- ranks_df$ENTREZID
gene_ranks <- sort(gene_ranks, decreasing = TRUE)
 
# Run a test GSEA with Hallmark
fg_test <- fgseaMultilevel(
  pathways = hallmark_sets,
  stats = gene_ranks,
  minSize = 5,
  maxSize = 500,
  scoreType = "pos",
  nproc = 1 #force single-threaded
)
 
fg_test %>% arrange(padj) %>% head()
```


#==============================================================
# 6. Run GSEA test for modules to check if all functions are working correctly
#==============================================================
```{r}
library(fgsea)
library(dplyr)
 
# Use scoreType = "pos" because kME values are all positive
score_type <- "pos"
 
# Loop through all modules
for (mod in modules) {
 
  cat("\n====================\n")
  cat("Module:", mod, "\n")
  # Genes in this module
  module_genes <- names(moduleColors[moduleColors == mod])
  # Only keep mapped genes
  ranks_df <- data.frame(
    SOMAmer = rownames(datKME),
    kME = datKME[, paste0("kME_", mod)]
  ) %>%
    filter(SOMAmer %in% module_genes) %>%
    inner_join(feature_map, by = "SOMAmer") %>%
    filter(is.finite(kME)) %>%
    distinct(ENTREZID, .keep_all = TRUE)
  cat("Mapped genes in module:", nrow(ranks_df), "\n")
  # Skip if too few genes
  if (nrow(ranks_df) < 15) {
    cat("Skipping module: less than 15 genes\n")
    next
  }
  # Build ranks
  gene_ranks <- ranks_df$kME
  names(gene_ranks) <- ranks_df$ENTREZID
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)
  # Run GSEA test for Hallmark (as example)
  fg_test <- fgseaMultilevel(
    pathways = hallmark_sets,
    stats = gene_ranks,
    minSize = 5,
    maxSize = 500,
    scoreType = score_type,
    nproc = 1 #force single-threaded
  )
  cat("Total pathways tested:", nrow(fg_test), "\n")
  cat("Significant pathways (padj < 0.05):", sum(fg_test$padj < 0.05), "\n")
  # Optional: print top 3 pathways by NES
  if(nrow(fg_test) > 0){
    print(fg_test %>% arrange(padj) %>% head(3))
  }
}
```


#==============================================================
# 7. Run GSEA Across All Modules
#==============================================================
```{r}
all_gsea_results <- list()
 
for (mod in modules) {
  cat("Processing module:", mod, "\n")
  h <- run_fgsea_module(mod, hallmark_sets)
  r <- run_fgsea_module(mod, reactome_sets)
  g <- run_fgsea_module(mod, go_sets)
 
  combined <- bind_rows(
    if (!is.null(h)) as_tibble(h) %>% mutate(Source = "Hallmark"),
    if (!is.null(r)) as_tibble(r) %>% mutate(Source = "Reactome"),
    if (!is.null(g)) as_tibble(g) %>% mutate(Source = "GO_BP")
  )
 
  if (!is.null(combined) && nrow(combined) > 0) {
    all_gsea_results[[mod]] <- combined
  } else {
    cat("Module", mod, "has no significant GSEA results\n")
  }
}

```


#==============================================================
# 8. Multi-Module Heatmap
#==============================================================
```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
 
# List of sources
sources <- c("Hallmark", "Reactome", "GO_BP")
 
for (src in sources) {
 
  # Combine results for this source
  combined_df <- bind_rows(
    lapply(names(all_gsea_results), function(m) {
      all_gsea_results[[m]] %>%
        dplyr::filter(Source == src, padj < 0.05) %>%
        dplyr::mutate(Module = m) %>%
        dplyr::select(pathway, NES, Module)
    })
  )
 
  if (nrow(combined_df) > 1) {
    heatmap_mat <- combined_df %>%
      distinct(pathway, Module, .keep_all = TRUE) %>%
      pivot_wider(names_from = Module, values_from = NES) %>%
      column_to_rownames("pathway") %>%
      mutate(across(everything(), ~replace_na(.x, 0)))
 
    # Auto-scale height: 0.3 inch per pathway row (adjust factor if needed)
    n_rows <- nrow(heatmap_mat)
    width_in <- max(12, ncol(heatmap_mat) * 1)         # 1 inch per module minimum 12
    height_in <- max(6, n_rows * 0.3)                 # 0.3 inch per pathway row
 
    # Save PDF for infinite resolution (recommended for large heatmaps)
    pdf(
      file = file.path(gsea_plots_dir, paste0("Heatmap_", src, ".pdf")),
      width = width_in,
      height = height_in
    )
 
    pheatmap(
      as.matrix(heatmap_mat),
      color = colorRampPalette(c("blue", "white", "red"))(100),
      main = paste("Significant Pathway NES Across Modules -", src),
      fontsize_row = 6,
      fontsize_col = 10,
      border_color = NA,
      clustering_method = "ward.D2"
    )
 
    dev.off()
    message("Saved heatmap for ", src, " with ", n_rows, " pathways.")
 
  } else {
    message("No significant pathways for source: ", src)

  }
}
```


#==============================================================
# 9. Dot Plot Function
#==============================================================
```{r}
library(stringr) # Essential for wrapping text
 
plot_fgsea_module_source <- function(df, module_name, source_name) {
  sig <- df %>% filter(Source == source_name, padj < 0.05)
  if (nrow(sig) == 0) return(NULL)
  # Pick top 10 by absolute NES
  top_sig <- sig %>% 
    slice_max(order_by = abs(NES), n = 10) %>%
    mutate(
      # 1. Clean up labels: remove prefixes like GO_BP_
      clean_pathway = gsub("^(GO_BP_|HALLMARK_|REACTOME_)", "", pathway),
      # 2. Replace underscores with spaces
      clean_pathway = gsub("_", " ", clean_pathway),
      # 3. Wrap long strings (limit to 40 characters per line)
      clean_pathway = str_wrap(clean_pathway, width = 40)
    )
  ggplot(top_sig, aes(
    x = NES,
    y = reorder(clean_pathway, NES),
    size = -log10(padj),
    color = NES
  )) +
    # Add thin vertical line at 0
    geom_vline(xintercept = 0, color = "grey70", linetype = "dashed") +
    geom_point(alpha = 0.8) +
    scale_color_gradient2(low = "#0571b0", mid = "#f7f7f7", high = "#ca0020") +
    theme_minimal(base_size = 12) +
    theme(
      # Adjust y-axis text to prevent overlap
      axis.text.y = element_text(lineheight = 0.8, color = "black"),
      axis.title.y = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "right",
      plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, size = 10)
    ) +
    labs(
      title = paste("Module:", module_name),
      subtitle = paste("Database:", source_name),
      x = "Normalized Enrichment Score (NES)",
      color = "Enrichment\n(NES)",
      size = "-log10(padj)"
    )
}
 
# Loop and Save
for (mod in names(all_gsea_results)) {
  for (src in sources) {
    p <- plot_fgsea_module_source(all_gsea_results[[mod]], mod, src)
    if (!is.null(p)) {
      # Path setup
      file_path <- file.path(gsea_plots_dir, paste0("DotPlot_", mod, "_", src, ".pdf"))
      # Use ggsave for better font handling
      # Fixed aspect ratio usually looks better for journals
      ggsave(
        filename = file_path,
        plot = p,
        width = 8, 
        height = 6, 
        device = "pdf",
        useDingbats = FALSE # Recommended for publication PDFs
      )
    }
  }
}
```


#==============================================================
# 10. CNEt plot for all modules with all databases that were conmsidered
#==============================================================
```{r}
library(enrichplot)
library(stringr)
 
plot_module_network_by_source <- function(df, module_name, source_name, n_pathways = 5) {
  # 1. Filter for specific Source and significance
  sig_res <- df %>% 
    filter(Source == source_name, padj < 0.05) %>% 
    slice_max(order_by = abs(NES), n = n_pathways)
  if(nrow(sig_res) == 0) return(NULL)
 
  # 2. Extract leading edge genes
  # fgsea stores these as a list of Entrez IDs
  gene_list <- sig_res$leadingEdge
  # Clean pathway names for better labeling in the network
  clean_names <- sig_res$pathway %>%
    gsub("^(GO_BP_|HALLMARK_|REACTOME_)", "", .) %>%
    gsub("_", " ", .) %>%
    str_wrap(width = 20)
  names(gene_list) <- clean_names
  # 3. Map Entrez IDs to Symbols for readability
  # Using your feature_map created earlier
  id_to_symbol <- feature_map$SYMBOL
  names(id_to_symbol) <- feature_map$ENTREZID
  gene_list_symbols <- lapply(gene_list, function(x) {
    symbols <- unname(id_to_symbol[as.character(x)])
    return(symbols[!is.na(symbols)]) # Remove any NAs
  })
 
# 4. Create the Network Plot
  p <- cnetplot(gene_list_symbols, 
                showCategory = n_pathways,
                layout = "circle", #options : gem, lgl, circle, defaut is "kk"
                #node_label = "group", 
                cex_label_gene = 0.7,      
                cex_label_category = 0.9,  
                colorEdge = TRUE,
                color_edge="category",
                shadowtext = "gene") +     
    labs(title = paste("Module:", module_name, "| Database:", source_name),
         subtitle = "Connections between top pathways and their leading-edge proteins") +
    theme(
      # Centering the title and subtitle
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5,margin = margin(b=30)),
      # Add margins to prevent the nodes/labels from being cut off at the edges
      # margin(top, right, bottom, left)
      plot.margin = margin(20, 20, 20, 20) 
    )
 
  return(p)
}
```





```{r}
sources <- c("Hallmark", "Reactome", "GO_BP")
 
for (mod in names(all_gsea_results)) {
  for (src in sources) {
    p_net <- plot_module_network_by_source(all_gsea_results[[mod]], mod, src)
    if (!is.null(p_net)) {
      # Create filename
      net_filename <- file.path(gsea_plots_dir, paste0("Network_", mod, "_", src, ".pdf"))
      # Save with a larger canvas to prevent label overlapping
      
      ggsave(
        filename = net_filename,
        plot = p_net,
        width = 8, 
        height = 6, 
        device = "pdf",
        useDingbats = FALSE # Recommended for publication PDFs
      )
      message("Saved network plot for Module: ", mod, " Source: ", src)
    }
  }
}
```


#==============================================================
# 11. Export Results 
#==============================================================
```{r}
wb <- createWorkbook()

for (mod in names(all_gsea_results)) {

  df <- all_gsea_results[[mod]] %>%
    filter(padj < 0.05) %>%
    arrange(Source, padj)

  if (nrow(df) > 0) {
    addWorksheet(wb, mod)
    writeData(wb, mod, df)
  }
}

saveWorkbook(
  wb,
  file.path(gsea_dir, "Combined_GSEA_Results.xlsx"),
  overwrite = TRUE
)

```


#==============================================================
# 11b. Database-specific GSEA results export
#==============================================================
```{r}
#==============================================================
# Save GSEA Results Separately Per Database
#==============================================================
 
library(openxlsx)
library(dplyr)
 
save_gsea_by_database <- function(all_gsea_results,
                                  outdir = gsea_dir,
                                  padj_cutoff = 0.05) {
  # Ensure output directory exists
  if (!dir.exists(outdir)) {
    dir.create(outdir, recursive = TRUE)
  }
  # Identify all unique databases
  all_sources <- unique(
    unlist(lapply(all_gsea_results, function(x) x$Source))
  )
  for (src in all_sources) {
    wb <- createWorkbook()
    cat("Creating workbook for:", src, "\n")
    for (mod in names(all_gsea_results)) {
      df <- all_gsea_results[[mod]] %>%
        filter(Source == src, padj < padj_cutoff) %>%
        arrange(padj)
      if (nrow(df) > 0) {
        addWorksheet(wb, mod)
        writeData(wb, mod, df)
      }
    }
    saveWorkbook(
      wb,
      file = file.path(outdir, paste0("GSEA_", src, "_Results.xlsx")),
      overwrite = TRUE
    )
  }
  cat("Finished exporting database-specific GSEA results.\n")
}


save_gsea_by_database(all_gsea_results)
```



#==============================================================
# 12. GSEA Running Score Plots (Mountain Plots)
#==============================================================
```{r}
library(fgsea)
library(ggplot2)
# Ensure directory exists

score_dir <- file.path(gsea_plots_dir, "GSEA_Scores")
dir.create(score_dir, recursive = TRUE, showWarnings = FALSE)
 
for (mod in names(all_gsea_results)) {

  # 1. Reconstruct the rank vector for the specific module
  kme_col <- paste0("kME_", mod)
  module_genes <- names(moduleColors[moduleColors == mod])

  ranks_df <- data.frame(
    SOMAmer = rownames(datKME),
    kME = datKME[, kme_col]

  ) %>%

    filter(SOMAmer %in% module_genes) %>%
    inner_join(feature_map, by = "SOMAmer") %>%
    filter(is.finite(kME)) %>%
    distinct(ENTREZID, .keep_all = TRUE)

  if(nrow(ranks_df) == 0) next
  gene_ranks <- ranks_df$kME
  names(gene_ranks) <- ranks_df$ENTREZID
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)

  # 2. Iterate through Hallmark, Reactome, and GO
  for (src in sources) {
    # Select the correct pathway list for the current source
    current_sets <- switch(src,
                           "Hallmark" = hallmark_sets,
                           "Reactome" = reactome_sets,
                           "GO_BP"    = go_sets)

    # Get top 2 significant pathways for this source
    top_paths <- all_gsea_results[[mod]] %>%
      filter(Source == src, padj < 0.05) %>%
      slice_min(order_by = padj, n = 2)

    if (nrow(top_paths) == 0) next

    for (i in 1:nrow(top_paths)) {
      path_name <- top_paths$pathway[i]
      stats <- top_paths[i, ] # Get stats for labeling

      # 3. Create the Running Score Plot
      p_mountain <- plotEnrichment(current_sets[[path_name]], gene_ranks) +
        labs(
          title = paste("Module:", mod, "-", src),
          subtitle = str_wrap(path_name, width = 60),
          caption = paste0("NES: ", round(stats$NES, 2), " | padj: ", format.pval(stats$padj, digits = 3))
        ) +
        
        theme_bw() +
        
        theme(
          plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
          plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b=10)),
          plot.caption = element_text(hjust = 0.5, size = 11, face = "italic", color = "darkred"),
          panel.grid.minor = element_blank(),
          plot.margin = margin(20, 20, 20, 20)
        )

      # 4. Save the plot
      # Use a safe filename
      clean_path_name <- str_trunc(gsub("[^[:alnum:]]", "_", path_name), 30)
      gsea_filename <- file.path(score_dir, paste0("Score_", mod, "_", src, "_", i, ".pdf"))

      ggsave(
        filename = gsea_filename,
        plot = p_mountain,
        width = 8,
        height = 6,
        device = "pdf",
        useDingbats = FALSE
      )
    }
  }
}
 
message("All GSEA Mountain plots have been saved to ", score_dir)
```

```{r}
#install.packages("report")
library(report)
report::report_packages()
```

