---
title: "Untitled"
author: "Raymond Otoo"
date: "2026-02-26"
output: pdf_document
---

#=====================================================================================
#
# 00- Set working directory
#=====================================================================================
```{r setup, include=FALSE}
source("project_setup.R")
paths <- init_project(knitr_root = TRUE)
supp_dir <- paths$supplementary_dir
getwd()
```


```{r}
library(WGCNA)
library(doParallel)
library(dplyr)
library(readxl)
library(writexl)
library(ggplot2)
library(ggrepel)
 
# Enable WGCNA internal multi-threading where applicable
enableWGCNAThreads()
 
# Set up Parallel Backend
n_cores_total <- parallel::detectCores()
# Use all cores minus 2 to keep the system stable
n_cores_to_use <- max(1, n_cores_total - 2) 
registerDoParallel(cores = n_cores_to_use)
 
message(paste("System has", n_cores_total, "cores. Parallel processing using:", n_cores_to_use))
```


#=====================================================================================
# 01- Module Preservation (Discovery & Validation) ADNI_as_Reference
#=====================================================================================
```{r}
load(file.path(paths$wgcna_dir, "03_output.RData"))
 
# 1.1 Subset Cohorts
sampleIDs <- rownames(expr)
cohort <- ifelse(grepl("^ADNI", sampleIDs), "ADNI", 
                 ifelse(grepl("^MAP", sampleIDs), "ADRC", NA))
 
expr_ADNI <- expr[cohort == "ADNI", ]
expr_ADRC <- expr[cohort == "ADRC", ]
 
# 1.2 Setup Multi-set Data
# We create a list containing both cohorts
multiExpr_Recip <- list(
  ADNI = list(data = expr_ADNI),
  ADRC = list(data = expr_ADRC)
)
 
# Use the module labels from your combined analysis
# We apply the same labels to both sets so we can test their preservation
multiColor_Recip <- list(
  ADNI = as.character(moduleColors),
  ADRC = as.character(moduleColors)
)
 
# 1.3 RUN 1: ADNI as Reference (Validate in ADRC)
message("Running Preservation: ADNI (Ref) -> ADRC (Test)")
pres <- modulePreservation(
  multiExpr_Recip,
  multiColor = multiColor_Recip,
  referenceNetworks = 1, # Index 1 is ADNI
  nPermutations = 500, # 500 is sufficient for robust Z-scores
  networkType       = "signed",
  randomSeed = 123,
  parallelCalculation = TRUE,
  verbose = 3
)
 

save(pres, file = file.path(supp_dir, "adni_as_ref_preservation.RData"))
```


#Extract Z Summary (ADNI → ADRC)
```{r}
pres$preservation$Z$ref.ADNI$inColumnsAlsoPresentIn.ADRC
```


#Z Summary Table (Key Modules Only)
```{r}
keyModules <- c("MEred", "MElightyellow", "MEblack", "MEmagenta")
keyModules_clean <- gsub("ME", "", keyModules)
 
# Identify ADRC slot (test network)
adrc_slot <- grep("ADRC",
                  names(pres$preservation$Z$ref.ADNI),
                  value = TRUE)
 
# Extract Z statistics
adrc_stats_Z <- pres$preservation$Z$ref.ADNI[[adrc_slot]][keyModules_clean, , drop=FALSE]
 
# Extract medianRank statistics
adrc_stats_median <- pres$preservation$observed$ref.ADNI[[adrc_slot]][keyModules_clean, , drop=FALSE]
 
# Build final table
final_z_summary <- data.frame(
  Module = rownames(adrc_stats_Z),
  Size = adrc_stats_Z$moduleSize,
  Z_ADRC = adrc_stats_Z$Zsummary.pres,
  medianRank_ADRC = adrc_stats_median$medianRank.pres,
  stringsAsFactors = FALSE
) %>%
  mutate(
    Preservation_Status = case_when(
      Z_ADRC > 10 ~ "Strongly Preserved",
      Z_ADRC > 2  ~ "Moderately Preserved",
      TRUE        ~ "Weak/No Preservation"
    )
  ) %>%
  arrange(medianRank_ADRC)   # lower medianRank = better preservation
 

print(final_z_summary)
```


#Save all for preservation results into one excel workbook
#NB: q statistics returned null so was omitted
```{r}
# Z statistics
df_Z <- pres$preservation$Z$ref.ADNI[[adrc_slot]][keyModules_clean, , drop=FALSE] %>%
  mutate(Module = rownames(.)) %>%
  relocate(Module)
 
# Observed statistics (contains medianRank)
df_observed <- pres$preservation$observed$ref.ADNI[[adrc_slot]][keyModules_clean, , drop=FALSE] %>%
  mutate(Module = rownames(.)) %>%
  relocate(Module)
 
# log.p statistics
df_logp <- pres$preservation$log.p$ref.ADNI[[adrc_slot]][keyModules_clean, , drop=FALSE] %>%
  mutate(Module = rownames(.)) %>%
  relocate(Module)
 
# q statistics
#null
 
# Write to single workbook
write_xlsx(
  list(
    Selected_ZSummary = final_z_summary,#built from table above
    Z_statistics = df_Z,
    Observed_statistics = df_observed,
    LogP_statistics = df_logp
    #Q_statistics = df_q
  ),
  path = file.path(supp_dir, "Module_Preservation_AllStats_ADNI_reference.xlsx")
)
 
cat("Excel workbook successfully written.")
```



#Preservation Plot (ADNI → ADRC)
```{r}
adrc_slot_name <- grep("ADRC",
                       names(pres$preservation$Z$ref.ADNI),
                       value = TRUE)

Z_ADRC <- pres$preservation$Z$ref.ADNI[[adrc_slot_name]][, "Zsummary.pres"]
moduleSize <- pres$preservation$Z$ref.ADNI[[adrc_slot_name]][, "moduleSize"]
moduleNames <- rownames(pres$preservation$Z$ref.ADNI[[adrc_slot_name]])

plot_data <- data.frame(
  size = moduleSize,
  z = Z_ADRC,
  label = moduleNames
) %>%
  filter(!label %in% c("gold", "grey")) %>%
  mutate(is_key = ifelse(label %in%
                           c("red","lightyellow","black","magenta"),
                         "Key","Other"))
```



```{r}
p_pres <- ggplot(plot_data, aes(x = size, y = z)) +
  geom_hline(yintercept = 2, linetype = "dashed",
             color = "blue", alpha = 0.7) +
  geom_hline(yintercept = 10, linetype = "dashed",
             color = "red", alpha = 0.7) +
  geom_point(aes(color = is_key, size = is_key), alpha = 0.8) +
  geom_text_repel(aes(label = label),
                  size = 3.5, max.overlaps = 15) +
  scale_color_manual(values = c("Key"="firebrick",
                                "Other"="grey40")) +
  scale_size_manual(values = c("Key"=4,
                               "Other"=2)) +
  labs(title = "Module Preservation: ADNI → ADRC",
       subtitle = "Zsummary Score vs Module Size",
       x = "Module Size",
       y = "Zsummary Score") +
  theme_minimal()
ggsave(file.path(supp_dir, "Figure1_Module_Preservation_ADNI_reference.png"),
       plot = p_pres, width = 8, height = 6, dpi = 600)
```



#=====================================================================================
#
# 02- STEP 2 — Cohort-Stratified Module–Trait Correlations (Sensitivity to AD Status and Adjusted Regression)
# Goal:
#   Show module–trait relationships are not driven by one cohort.
#  Goal: Isolate the CSVD signal from Alzheimer's pathology.Interpretation: By adding AD_status as a covariate, the resulting p-value for the Trait tells us if the protein module is specifically related to vascular issues (CSVD) regardless of whether the      #   patient has AD.
#=====================================================================================
```{r}
load(file.path(paths$wgcna_dir, "WGCNA2_core_outputs.RData"))
 
# Function to run AD-adjusted regressions for all key modules
# Addresses Point 5: MEmodule ~ trait + age + sex + AD_status
run_adjusted_models <- function(MEs, traits, module_name, trait_name) {
  df_reg <- data.frame(
    ME = MEs[[module_name]],
    Trait = traits[[trait_name]],
    Age = traits$Age_at_draw,
    Sex = traits$Sex,
    AD = traits$AD_status
  ) %>% na.omit()
  fit <- lm(ME ~ Trait + Age + Sex + AD, data = df_reg)
  res <- coef(summary(fit))["Trait", ]
  return(data.frame(Module = module_name, Trait = trait_name, 
                    Beta = res[1], SE = res[2], P = res[4]))
}
 
# Mapping modules to their primary traits of interest
analysis_map <- list(
  list(m = "MEred", t = "fsrp"),
  list(m = "MEred", t = "htnscore"),
  list(m = "MElightyellow", t = "PC1"),
  list(m = "MEblack", t = "caa"),
  list(m = "MEmagenta", t = "bmi")
)
 
# Execute via purrr or lapply
adj_results <- lapply(analysis_map, function(x) {
  run_adjusted_models(MEs[rownames(expr),], traits_interest[rownames(expr),], x$m, x$t)
}) %>% bind_rows()
 
print(adj_results)
```



#=====================================================================================
#
# 03-Module Membership (kME Extraction) 
# Orthogonal Evidence: kME Extraction
#=====================================================================================
```{r}
# Addresses Point 1: Quantify how much proteins in key modules align with trait hits
kME_table <- signedKME(expr, MEs)
 
# Create a clean table for export
final_protein_kME <- data.frame(
  ProteinID = colnames(expr),
  ModuleColor = moduleColors
) %>%
  cbind(kME_table)
 
# Export to Excel for reviewers
write_xlsx(final_protein_kME, file.path(supp_dir, "Protein_Module_Membership_kME.xlsx"))
```


#=====================================================================================
# Point 6: Coefficient Comparison (Final Evidence)
# Goal: Quantify the shift in association after adjustment.
# Interpretation: If the Beta (slope) remains similar to the unadjusted correlation, the association is "Orthogonal" to AD status. This is your strongest argument for novelty in CSVD subtyping.
#=====================================================================================
```{r}
sampleIDs <- rownames(expr)
 
cohort <- ifelse(grepl("^ADNI", sampleIDs), "ADNI",
          ifelse(grepl("^MAP", sampleIDs), "ADRC", NA))
```


#Split MEs and traits
```{r}
MEs_ordered <- MEs[rownames(expr), ]
traits_ordered <- traits_interest[rownames(expr), ]
 
MEs_ADNI  <- MEs_ordered[cohort == "ADNI", ]
MEs_ADRC  <- MEs_ordered[cohort == "ADRC", ]
 
traits_ADNI <- traits_ordered[cohort == "ADNI", ]
traits_ADRC <- traits_ordered[cohort == "ADRC", ]
```



```{r}
# Compare raw correlation vs adjusted Beta
comparison_table <- adj_results %>%
  mutate(Original_Cor = sapply(1:n(), function(i) {
    cor(MEs_ordered[[Module[i]]], traits_interest[[Trait[i]]], use="p")
  }))
comparison_table

write_xlsx(comparison_table, file.path(supp_dir, "Final_Trait_Sensitivity_Summary.xlsx"))
```


#=====================================================================================
# Orthogonal Evidence: Convergence with Single-Protein Hits
# Goal: Verify that the "Module" findings match "Single Protein" findings from earlier in your project.Interpretation: If the proteins with the highest kME in the red module are the same proteins that showed up as top hits in your Olink/SOMA volcano plots for   # FSRP, you have "convergent evidence" from two different statistical methods (WGCNA and Linear Regression).
#=====================================================================================
```{r}
# Check convergence (Example for Red module)
red_hub_proteins <- final_protein_kME %>% filter(moduleColors == "red") %>% arrange(desc(kMEred))
head(red_hub_proteins, 10)
```


#3. Sensitivity Forest Plot (Points 4, 5 & 6)
#This plot visualizes how the association strength (Beta) changes between the Unadjusted model and the AD-Adjusted model. It proves your findings are independent of Alzheimer's pathology.
```{r}
# Prepare data for plotting from comparison_table
forest_data <- comparison_table %>%
  dplyr::select(Module, Trait, Unadj_Beta = Original_Cor, Adj_Beta = Beta, SE) %>%
  tidyr::pivot_longer(cols = c(Unadj_Beta, Adj_Beta), names_to = "Model", values_to = "Estimate")
 
ggplot(forest_data, aes(x = Estimate, y = interaction(Module, Trait), color = Model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbarh(aes(xmin = Estimate - (1.96 * SE), xmax = Estimate + (1.96 * SE)), 
                 position = position_dodge(width = 0.5), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  labs(title = "Sensitivity Analysis: AD Adjustment",
       subtitle = "Error bars represent 95% CI",
       y = "Module-Trait Pair", x = "Effect Size (Beta/Correlation)") +
  theme_bw()
ggsave(file.path(supp_dir, "Figure3_AD_Sensitivity_Forest.png"), width = 8, height = 6, dpi = 600)
```


#4. Top Hub Proteins (kME) Lollipop plot  --- Convergence
#This plot identifies the "leading edge" proteins in your key modules. It is the "orthogonal evidence" showing that the modules are driven by biologically relevant hub proteins.
```{r}
# Select top 5 proteins per key module
top_hubs <- final_protein_kME %>%
  filter(ModuleColor %in% c("red", "lightyellow", "black", "magenta")) %>%
  tidyr::pivot_longer(cols = starts_with("kME"), names_to = "kME_Type", values_to = "kME_Value") %>%
  # Match the kME column to the module color
  filter(gsub("kME", "", kME_Type) == ModuleColor) %>%
  group_by(ModuleColor) %>%
  slice_max(order_by = kME_Value, n = 10)
 
ggplot(top_hubs, aes(x = reorder(ProteinID, kME_Value), y = kME_Value)) +
  geom_segment(aes(xend = reorder(ProteinID, kME_Value), yend = 0), color = "black") +
  geom_point(aes(color = ModuleColor), size = 4) +
  coord_flip() +
  facet_wrap(~ModuleColor, scales = "free_y") +
  scale_color_identity() +
  labs(title = "Key Module Hub Proteins",
       subtitle = "Ranked by Module Membership (kME)",
       x = "Protein (SOMA ID / Gene)", y = "kME Strength") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank()) # Removes horizontal noise
ggsave(file.path(supp_dir, "Figure4_Top_Hub_Proteins.png"), width = 10, height = 8, dpi = 600)
```


#2.3 Corrlelations
```{r}
cor_ADNI <- cor(MEs_ADNI, traits_ADNI, use="pairwise.complete.obs")
p_ADNI   <- corPvalueStudent(cor_ADNI, nrow(MEs_ADNI))
 
cor_ADRC <- cor(MEs_ADRC, traits_ADRC, use="pairwise.complete.obs")
p_ADRC   <- corPvalueStudent(cor_ADRC, nrow(MEs_ADRC))
```


```{r}
keyModules <- c("MEred", "MElightyellow", "MEblack", "MEmagenta")
 
cor_ADNI[keyModules, ]
cor_ADRC[keyModules, ]
```



```{r}
# Focus only on your keyModules for clarity
key_cor_ADNI <- cor_ADNI[keyModules, , drop=FALSE]
key_p_ADNI   <- p_ADNI[keyModules, , drop=FALSE]
 
key_cor_ADRC <- cor_ADRC[keyModules, , drop=FALSE]
key_p_ADRC   <- p_ADRC[keyModules, , drop=FALSE]
```



##Module_Trait_Comparison_vertical ADNI-ADRC
```{r}
plot_wgcna_heatmap <- function(cor_matrix, p_matrix, title) {
  textMatrix <- paste(signif(cor_matrix, 2), "\n(",
                      signif(p_matrix, 1), ")", sep = "")
  dim(textMatrix) <- dim(cor_matrix)
  par(mar = c(10, 12, 4, 4))
  labeledHeatmap(
    Matrix = cor_matrix,
    xLabels = colnames(cor_matrix),
    yLabels = rownames(cor_matrix),
    ySymbols = rownames(cor_matrix),
    colorLabels = FALSE,
    colors = blueWhiteRed(50),
    textMatrix = textMatrix,
    setStdMargins = FALSE,
    cex.text = 0.6,
    cex.lab.x = 0.8,
    cex.lab.y = 0.8,
    zlim = c(-1,1),
    main = title
  )
}
 
png(file.path(supp_dir, "Module_Trait_Comparison_vertical.png"), width = 8000, height = 7000, res = 600)
 
par(mfrow = c(2, 1))  # 2 rows, 1 column
 
plot_wgcna_heatmap(key_cor_ADNI, key_p_ADNI,
                   "ADNI: Module–Trait Relationships")
 
plot_wgcna_heatmap(key_cor_ADRC, key_p_ADRC,
                   "ADRC: Module–Trait Relationships")
 
dev.off()
```
